---
title: "Figures"
author: "Summer Heschong"
date: "2025-03-05"
output: html_document
---
#Setup 
```{r setup}
library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(moments)

#load data
wetland_data <- read.csv(here('Data/Processed/WQ-Data-NCDWR-2005-to-2015-NCSU-2015-to-2019.csv'))

```

#Inspect and Wrangle data
```{r wrangle}

#I tried to make some graphs and I think there might not be enough data for sites 1014 and 1018

#filter for just NWCA project sites
NWCA <- wetland_data %>%
  filter(Project == 'NWCA Intensification')

#see which sites in the NWCA project have the most observations
sort(table(NWCA$Site), decreasing = TRUE)

#lets try 1161 and 1012 instead? 

#ok we might have to pick sites that aren't in the NWCA project. There's a lot less data than it seems because all the water quality parameters are grouped in one column

#which sites overall have the most observations?
sort(table(wetland_data$Site), decreasing = TRUE)

#Hog Farm Upper, Walmart, and Nahunta are all in project Headwater and they have up and downstream data

#create a dataset for each site
HogFarm <- wetland_data %>%
  filter(Site == 'Hog Farm Upper')

Walmart <- wetland_data %>%
  filter(Site == 'Walmart')

Nahunta <- wetland_data %>%
  filter(Site == 'Nahunta')

#create a dataset with just our sites
Project_sites <- wetland_data %>%
  filter(Site %in% c('Hog Farm Upper', 'Walmart', 'Nahunta'))
```

#Visualize data
##Water quality upstream vs downstream
```{r visualize up and dwn}
#Compare water quality parameters upstream and downstream for same site

#First filter dataset for one parameter
HogFarm_DO <- HogFarm %>%
  filter(Parameter == 'Dissolved Oxygen (%)') %>%
  mutate(Location.Type = case_when(
    Location.Type == "Further Downstream"~"Downstream",
    TRUE ~ Location.Type))

#create histograms of parameter by location
HogFarm_DO_hist <- ggplot(HogFarm_DO, aes(x = Result, fill = Location.Type)) +
  geom_histogram(bins = 6, binwidth = 12, color = 'black') +
  facet_wrap(~Location.Type) +
  labs(title = "Dissolved Oxygen Higher Downstream than Upstream in Coastal Wetland",
       caption = "Histogram comparing Dissolved Oxygen (%) \n upstream vs. downstream at a coastal plain wetland in Samson County, NC",
       x = "Dissolved Oxygen (%)",
       fill = "Location") +
  theme(plot.caption = element_text(hjust = 0.5),
        plot.title = element_text(hjust = 0.5))

HogFarm_DO_hist

#save plot as .jpeg
ggsave(HogFarm_DO_hist, filename="HogFarm_DO_Fig1.jpg", 
       height=20, width=20, units="cm")

#create scatterplot of parameter by location
HogFarm_ph_scatter <- ggplot(HogFarm_pH, aes(x = Location.Type, y = Result)) +
  geom_point(size = 3, alpha = 0.5)
HogFarm_ph_scatter

#create boxplot
HogFarm_DO_box <- ggplot(HogFarm_DO, aes(x = Location.Type, y = Result, 
                                         fill = Location.Type)) +
  geom_boxplot() +
  labs(title = "Dissolved Oxygen Higher Downstream than Upstream in Coastal Wetland",
       caption = "Dissolved Oxygen (%) upstream vs. downstream at a coastal plain wetland in Sampson County, NC",
       x = "Location",
       y = "Dissolved Oxygen (%)") +
  theme(legend.position = "none",
        plot.caption = element_text(hjust = 0.5),
        plot.title = element_text(hjust = 0.5))

HogFarm_DO_box

#save plot
ggsave(HogFarm_DO_box, filename="HogFarm_DO_Fig2.jpg", 
       height=20, width=20, units="cm")

#Should we combine Downstream and Further Downstream?
```

## Metal Concentrations and pH
```{r visualize metal}
#first filter data for pH and copper (can change for other metals) and separate them into 2 columns

metal_vs_pH <- Project_sites %>%
  select(Station.Name, Parameter, Result) %>%
  filter(Parameter == 'Copper' | Parameter == 'pH') %>%
  pivot_wider(names_from = Parameter, values_from = Result, values_fn = mean)
#there were duplicate values when i first did pivot_wider so i used the mean of the duplicates. Idk how much this matters statistically? 

#create a scatterplot to examine relationship between metal concentrations and pH
ggplot(metal_vs_pH, aes(x = pH, y = Copper)) +
  geom_point() +
  geom_smooth(method = lm)

#I don't think this plot really works?
```

# If lead concentrations at each site meet standards (one-sample t-test)

```{r TBD}

```

# How lead concetrations differ by reach in hogfarm

```{r Create Filtered Datasets}
#Note to Summer and Nicole: Once I learn how, I would like to try a two-way ANOVA for both site and reach. What do you guys think? Is our end goal to just stick to one site?

Hogfarm_upstream_pb <- HogFarm %>%
  filter(Location.Type=="Upstream",
         Parameter=="Lead")

Hogfarm_downstream_pb <- HogFarm %>%
    filter(Location.Type=="Downstream",
         Parameter=="Lead")

Hogfarm_further_downstream_pb <- HogFarm %>%
    filter(Location.Type=="Further Downstream",
         Parameter=="Lead")

#Further Downstream has a sample size of 11 for lead lmao
```

``` {r Skewness and Kurtosis}

#Calculate skewness and kurtosis for Upstream

skewness(Hogfarm_upstream_pb$Result, na.rm = TRUE)

kurtosis(Hogfarm_upstream_pb$Result, na.rm = TRUE)

#Calculate skewness and kurtosis for Downstream

skewness(Hogfarm_downstream_pb$Result, na.rm = TRUE)

kurtosis(Hogfarm_downstream_pb$Result, na.rm = TRUE)

#Calculate skewness and kurtosis for Further Downstream

skewness(Hogfarm_further_downstream_pb$Result, na.rm = TRUE)

kurtosis(Hogfarm_further_downstream_pb$Result, na.rm = TRUE)



```

I am going to go ahead and say it is safe to log transform these values. Fingers crossed!

```{r Log Transform}

#Make new log datasets
Hogfarm_Upstream_log <- Hogfarm_upstream_pb %>%
  mutate(Result=log10(Result))

Hogfarm_Downstream_log <- Hogfarm_downstream_pb %>%
  mutate(Result=log10(Result))

Hogfarm_Further_Downstream_log <- Hogfarm_further_downstream_pb %>%
  mutate(Result=log10(Result))

#Check histogram of each before even checking skewness or kurtosis

ggplot(Hogfarm_Upstream_log, aes(x=Result)) +
  geom_histogram()

ggplot(Hogfarm_Downstream_log, aes(x=Result)) +
  geom_histogram()

ggplot(Hogfarm_Further_Downstream_log, aes(x=Result)) +
  geom_histogram()


#Yeah I am gonna do a KW test. Tbh this test will likely not get used but a) It's a start and b) Let's just see what kind of data we get
```

```{r Kruskal-Wallis}

#Filter whole hogfarm dataset for just lead

Hogfarm_Lead <- HogFarm %>%
  filter(Parameter=="Lead")

kruskal_test <- kruskal.test(Result ~ Location.Type, data = Hogfarm_Lead)

kruskal_test #Display result
```


Ok so with that 0.31 p-value, I am just gonna make the figure lol.

```{r Communicating Hogfarm Lead Results, fig.cap="Distributions of lead concentrations (ppb) in upstream, downstream, and father downstream of Hogfarm headwaters in the Coastal Plains."}
#Calculate median and SD
Hogfarm_Lead_Summary <- Hogfarm_Lead %>%
  group_by(Location.Type) %>%
  summarize(mean=mean(Result, na.rm = TRUE),
            sd=sd(Result, na.rm = TRUE)) %>%
  ungroup()

Hogfarm_Lead_Summary <- Hogfarm_Lead_Summary %>%
  rename(Result = mean)

Hogfarm_Lead_Fig <- ggplot() +
  geom_jitter(data = Hogfarm_Lead,
              aes(x=Location.Type,
                  y=Result,
                  color=Location.Type),
              alpha=0.5,
              size=0.5) +
  geom_point(data=Hogfarm_Lead_Summary,
             aes(x=Location.Type, 
                 y=Result,
                 color=Location.Type),
             size=3) +
  geom_errorbar(data=Hogfarm_Lead_Summary,
                aes(x=Location.Type,
                    ymin=Result-sd,
                    ymax=Result+sd,
                    color=Location.Type),
                width=0.10, size=1) +
  annotate("text", x="Upstream", y=150, label="a", size=8) +
  annotate("text", x="Downstream", y=150, label="a", size=8) +
  annotate("text", x="Further Downstream", y=150, label="a", size=8) +
  scale_color_manual(values=c("darkorange","purple","cyan4")) +
  labs(x="Reach",
       y="Pb Concentrations (ppb)", caption = "Figure x: Distributions of lead concentrations (ppb) in upstream, downstream, and \n father downstream of Hogfarm headwaters in the Coastal Plains.") +
  theme_bw() +
  theme(legend.position = "none", plot.caption = element_text(hjust = 0.5))

Hogfarm_Lead_Fig

ggsave(Hogfarm_Lead_Fig, filename="Hogfarm_Lead_Fig.jpg", height=20, width=20, units="cm")


```

