---
title: "Ammonia Model"
author: "Summer Heschong"
date: "2025-03-31"
output: pdf_document
---
# Setup 
```{r setup}
#load packages
library(here)
library(tidyverse)
library(nlme)

wetlands <- read.csv(here('Data/Processed/Combined_Data_NArm.csv'))

```

# Step 1 - Research Question: What predicts ammonium levels in wetlands

# Step 2 - Examine Data
## Display raw counts and distributions of data
```{r histograms}

#remove outliers using Nicole's code
wetlands <- wetlands %>%
  mutate(across(where(is.numeric), 
                ~ ifelse(abs
                         (. - mean
                         (., na.rm = TRUE)) > 3 * sd
                         (., na.rm = TRUE), NA, .)))

#create histograms
ggplot(wetlands, aes(x = Temp_C)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = SpCond_mScm)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Cond_mScm)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = TDS_mgl)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Sal_ppt)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = DO_mgL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = pH)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = TSS_mgL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = fecal_coliform_colonies_per100mL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Unfiltered_TP_ugL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Site)) +
geom_bar()

ggplot(wetlands, aes(x = Season)) +
geom_bar()
```

## Display relationships between predictor variables and outcome variable
```{r scatter}
#create scatterplots
ggplot(wetlands, aes(x = Temp_C, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = SpCond_mScm, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = TDS_mgl, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = Sal_ppt, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = DO_mgL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = pH, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = TSS_mgL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = fecal_coliform_colonies_per100mL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = Unfiltered_TP_ugL, y = Filtered_NHx_ugL)) +
geom_point()

```

## everything needs to be log transformed?
```{r log transform}
#log transform everything except DO
log_wetland <- wetlands %>%
  mutate(across(c(SpCond_mScm, Cond_mScm, 
                  TDS_mgl, Sal_ppt, 
                  Filtered_NHx_ugL, TSS_mgL, 
                  Unfiltered_TP_ugL, Filtered_NOx_ugL,
                  Unfiltered_TN_ugL), 
                ~ log(.+1)))
#change column names
log_transformed <- log_wetland %>% 
  rename(Log_SpCond_mScm = SpCond_mScm, 
         Log_Cond_mScm = Cond_mScm, 
         Log_TDS_mgl = TDS_mgl ,   
         Log_Sal_ppt = Sal_ppt, 
         Log_Unfiltered_TN_ugL = Unfiltered_TN_ugL,
         Log_Filtered_NOx_ugL = Filtered_NOx_ugL,
         Log_Filtered_NHx_ugL = Filtered_NHx_ugL,
         Log_Unfiltered_TP_ugL = Unfiltered_TP_ugL,
         Log_TSS_mgL = TSS_mgL)

#write.csv(log_transformed, "Data/Processed/Log_Transformed_Data.csv", row.names = FALSE)

#re-vistualize data

ggplot(log_wetland, aes(x = Temp_C, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = SpCond_mScm, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = TDS_mgl, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = Sal_ppt, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = DO_mgL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = pH, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = TSS_mgL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = fecal_coliform_colonies_per100mL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(log_wetland, aes(x = Unfiltered_TP_ugL, y = Filtered_NHx_ugL)) +
geom_point()

```

## Display fixed effect variables across random effect variable
```{r boxplot}
#create boxplots of each fixed effect variable

ggplot(log_wetland, aes(x = Site, y = Temp_C)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = SpCond_mScm)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = TDS_mgl)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Sal_ppt)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = DO_mgL)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = pH)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Unfiltered_TP_ugL)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Filtered_NHx_ugL)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = TSS_mgL)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = fecal_coliform_colonies_per100mL)) +
geom_boxplot()
```

## Check for correlation between variables
```{r correlation}

#Temp vs: 
cor.test(log_wetland$Temp_C, log_wetland$SpCond_mScm)
cor.test(log_wetland$Temp_C, log_wetland$TDS_mgl)
cor.test(log_wetland$Temp_C, log_wetland$Sal_ppt)
cor.test(log_wetland$Temp_C, log_wetland$DO_mgL)
cor.test(log_wetland$Temp_C, log_wetland$pH)
cor.test(log_wetland$Temp_C, log_wetland$Unfiltered_TP_ugL)
cor.test(log_wetland$Temp_C, log_wetland$TSS_mgL)
cor.test(log_wetland$Temp_C, log_wetland$fecal_coliform_colonies_per100mL)

#SpCond vs:
cor.test(log_wetland$SpCond_mScm, log_wetland$TDS_mgl)
cor.test(log_wetland$SpCond_mScm, log_wetland$Sal_ppt)
cor.test(log_wetland$SpCond_mScm, log_wetland$DO_mgL)
cor.test(log_wetland$SpCond_mScm, log_wetland$pH)
cor.test(log_wetland$SpCond_mScm, log_wetland$Unfiltered_TP_ugL)
cor.test(log_wetland$SpCond_mScm, log_wetland$TSS_mgL)
cor.test(log_wetland$SpCond_mScm, log_wetland$fecal_coliform_colonies_per100mL)

#TDS vs:
cor.test(log_wetland$TDS_mgl, log_wetland$Sal_ppt)
cor.test(log_wetland$TDS_mgl, log_wetland$DO_mgL)
cor.test(log_wetland$TDS_mgl, log_wetland$pH)
cor.test(log_wetland$TDS_mgl, log_wetland$Unfiltered_TP_ugL)
cor.test(log_wetland$TDS_mgl, log_wetland$TSS_mgL)
cor.test(log_wetland$TDS_mgl, log_wetland$fecal_coliform_colonies_per100mL)

#Sal vs
cor.test(log_wetland$Sal_ppt, log_wetland$DO_mgL)
cor.test(log_wetland$Sal_ppt, log_wetland$pH)
cor.test(log_wetland$Sal_ppt, log_wetland$Unfiltered_TP_ugL)
cor.test(log_wetland$Sal_ppt, log_wetland$TSS_mgL)
cor.test(log_wetland$Sal_ppt, log_wetland$fecal_coliform_colonies_per100mL)

#DO vs
cor.test(log_wetland$DO_mgL, log_wetland$pH)
cor.test(log_wetland$DO_mgL, log_wetland$Unfiltered_TP_ugL)
cor.test(log_wetland$DO_mgL, log_wetland$TSS_mgL)
cor.test(log_wetland$DO_mgL, log_wetland$fecal_coliform_colonies_per100mL)
#pH vs
cor.test(log_wetland$pH, log_wetland$Unfiltered_TP_ugL)
cor.test(log_wetland$pH, log_wetland$TSS_mgL)
cor.test(log_wetland$pH, log_wetland$fecal_coliform_colonies_per100mL)
#TP vs
cor.test(log_wetland$Unfiltered_TP_ugL, log_wetland$TSS_mgL)
cor.test(log_wetland$Unfiltered_TP_ugL, log_wetland$fecal_coliform_colonies_per100mL)

#TSS vs
cor.test(log_wetland$TSS_mgL, log_wetland$fecal_coliform_colonies_per100mL)

#Predictor and Outcome Variable correlations 
cor.test(log_wetland$Temp_C, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$SpCond_mScm, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$TDS_mgl, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$Sal_ppt, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$DO_mgL, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$pH, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$Unfiltered_TP_ugL, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$TSS_mgL, log_wetland$Filtered_NHx_ugL)
cor.test(log_wetland$fecal_coliform_colonies_per100mL, log_wetland$Filtered_NHx_ugL)


```
SpCond and Sal are too correlated to use together - makes sense. Also all the predictor variables have very low correlations with the outcome variable :( 

# Step 3 - Fit regular regression model
I think I'm going to try three models.
```{r generalized regression models}
#first convert Season to a factor
log_wetland$Season <- factor(log_wetland$Season, 
                             levels = c('Winter', 'Spring', 'Summer', 'Fall'))
# fit regression models
mod1 <- lm(Filtered_NHx_ugL ~ Temp_C + pH + Season,
            data = log_wetland)

mod2 <- lm(Filtered_NHx_ugL ~ Temp_C + DO_mgL + Season,
            data = log_wetland)

mod3 <- lm(Filtered_NHx_ugL ~ Unfiltered_TP_ugL + DO_mgL + Temp_C + pH + Season,
            data = log_wetland)

#examine model outputs and residuals
summary(mod1)
plot(mod1)

summary(mod2)
plot(mod2)

summary(mod3)
plot(mod3)

```

# Step 4 - Fit Mixed Effects Model and Examine Model Results
## use gls model first 
```{r gls}
#remove NA's
log_wetland <- na.omit(log_wetland)

#first refit models using generalized least squares
GLS1 <- gls(Filtered_NHx_ugL ~ Temp_C + pH + Season,
            data = log_wetland)
GLS2 <- gls(Filtered_NHx_ugL ~ Temp_C + DO_mgL + Season,
            data = log_wetland)
GLS3 <- gls(Filtered_NHx_ugL ~ Temp_C + DO_mgL + Unfiltered_TP_ugL + pH + Season,
            data = log_wetland)

#look at model outputs
summary(GLS1)
plot(GLS1)

summary(GLS2)
plot(GLS2)

summary(GLS3)
plot(GLS3)
```
Looking better than before yay

##now fit to mixed effects model
```{r mixed}
#fit data to mixed effects models
MEM1 <- lme(Filtered_NHx_ugL ~ Temp_C + pH + Season,
            random = ~ 1|Site, data = log_wetland)
MEM2 <- lme(Filtered_NHx_ugL ~ Temp_C + DO_mgL + Season,
            random = ~ 1|Site, data = log_wetland)
MEM3 <- lme(Filtered_NHx_ugL ~ Temp_C + DO_mgL + Unfiltered_TP_ugL + pH + Season,
            random = ~ 1|Site, data = log_wetland)

#Compare AIC values between gls and lme models
AIC(GLS1, MEM1)
AIC(GLS2, MEM2)
AIC(GLS3, MEM3)

#Examine models residual plots
plot(MEM1)
qqnorm(MEM1)

plot(MEM2)
qqnorm(MEM2)

plot(MEM3)
qqnorm(MEM3)

#Examine results
summary(MEM1)
summary(MEM2)
summary(MEM3)

```
The mixed effect models have lower AIC values than the gls linear regression models


