---
title: "Ammonia Model"
author: "Summer Heschong"
date: "2025-03-31"
output:
  pdf_document: default
  html_document:
    df_print: paged
---
# Setup 
```{r setup}
#load packages
library(here)
library(tidyverse)
library(nlme)
library(gtsummary)
library(cowplot)

wetlands <- read.csv(here('Data/Processed/Combined_Data_NArm.csv'))

```

# Step 1 - Research Question: What predicts ammonium levels in wetlands

# Step 2 - Examine Data
## Display raw counts and distributions of data
```{r histograms}

#remove outliers using Nicole's code
wetlands <- wetlands %>%
  mutate(across(where(is.numeric), 
                ~ ifelse(abs
                         (. - mean
                         (., na.rm = TRUE)) > 3 * sd
                         (., na.rm = TRUE), NA, .)))

#create histograms
ggplot(wetlands, aes(x = SpCond_mScm)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Cond_mScm)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = TDS_mgl)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Sal_ppt)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = DO_mgL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = pH)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = TSS_mgL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Unfiltered_TP_ugL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Filtered_NHx_ugL)) +
geom_histogram(color = 'black')

ggplot(wetlands, aes(x = Site)) +
geom_bar()

ggplot(wetlands, aes(x = Season)) +
geom_bar()
```

## Display relationships between predictor variables and outcome variable
```{r scatter}
#create scatterplots
ggplot(wetlands, aes(x = SpCond_mScm, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = TDS_mgl, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = Sal_ppt, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = DO_mgL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = pH, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = TSS_mgL, y = Filtered_NHx_ugL)) +
geom_point()

ggplot(wetlands, aes(x = Unfiltered_TP_ugL, y = Filtered_NHx_ugL)) +
geom_point()

```

## everything needs to be log transformed? (save log-transformed data as new dataset)
```{r log transform}
#log transform everything except DO, and pH
#log_wetland <- wetlands %>%
 # mutate(across(c(SpCond_mScm, Cond_mScm, 
               #   TDS_mgl, Sal_ppt, 
                 # Filtered_NHx_ugL, TSS_mgL, 
                 # Unfiltered_TP_ugL, Filtered_NOx_ugL,
                 # Unfiltered_TN_ugL), 
             #   ~ log(.+1)))

#change column names
#log_transformed <- log_wetland %>% 
 # rename(Log_SpCond_mScm = SpCond_mScm, 
  #       Log_Cond_mScm = Cond_mScm, 
   #      Log_TDS_mgl = TDS_mgl ,   
    #     Log_Sal_ppt = Sal_ppt, 
     #    Log_Unfiltered_TN_ugL = Unfiltered_TN_ugL,
      #   Log_Filtered_NOx_ugL = Filtered_NOx_ugL,
       #  Log_Filtered_NHx_ugL = Filtered_NHx_ugL,
        # Log_Unfiltered_TP_ugL = Unfiltered_TP_ugL,
         #Log_TSS_mgL = TSS_mgL)

#save as new dataset
#write.csv(log_transformed, "Data/Processed/Log_Transformed_Data.csv", row.names = FALSE)
```

## re-visualize data
```{r revis}
#load new dataset
log_wetland <- read.csv(here('Data/Processed/Log_Transformed_Data.csv'))

#create histograms of all log-transformed variables
ggplot(log_wetland, aes(x = Log_SpCond_mScm)) +
geom_histogram(color = 'black')

ggplot(log_wetland, aes(x = Log_Cond_mScm)) +
geom_histogram(color = 'black') 

ggplot(log_wetland, aes(x = Log_TDS_mgl)) +
geom_histogram(color = 'black') 

ggplot(log_wetland, aes(x = Log_Sal_ppt)) +
geom_histogram(color = 'black')

ggplot(log_wetland, aes(x = Log_TSS_mgL)) +
geom_histogram(color = 'black')

ggplot(log_wetland, aes(x = Log_Unfiltered_TP_ugL)) +
geom_histogram(color = 'black') 

ggplot(log_wetland, aes(x = Log_Filtered_NHx_ugL)) +
geom_histogram(color = 'black') 

#create scatterplots of predictor vs outcome var
ggplot(log_wetland, aes(x = Log_SpCond_mScm, y = Log_Filtered_NHx_ugL)) +
geom_point() +
geom_smooth(method = "lm")

ggplot(log_wetland, aes(x = Log_TDS_mgl, y = Log_Filtered_NHx_ugL)) +
geom_point() + 
  geom_smooth(method = "lm")

ggplot(log_wetland, aes(x = Log_Sal_ppt, y = Log_Filtered_NHx_ugL)) +
geom_point() +
geom_smooth(method = "lm")

ggplot(log_wetland, aes(x = DO_mgL, y = Log_Filtered_NHx_ugL)) +
geom_point() +
geom_smooth(method = "lm")

ggplot(log_wetland, aes(x = pH, y = Log_Filtered_NHx_ugL)) +
geom_point() +
geom_smooth(method = "lm")

ggplot(log_wetland, aes(x = Log_TSS_mgL, y = Log_Filtered_NHx_ugL)) +
geom_point() +
geom_smooth(method = "lm")

ggplot(log_wetland, aes(x = Log_Unfiltered_TP_ugL, y = Log_Filtered_NHx_ugL)) +
geom_point() +
geom_smooth(method = "lm")

```

## Display fixed effect variables across random effect variable
```{r boxplot}
#create boxplots of each fixed effect variable
ggplot(log_wetland, aes(x = Site, y = Log_SpCond_mScm)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Log_TDS_mgl)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Log_Sal_ppt)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = DO_mgL)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = pH)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Log_Unfiltered_TP_ugL)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Log_Filtered_NHx_ugL)) +
geom_boxplot()

ggplot(log_wetland, aes(x = Site, y = Log_TSS_mgL)) +
geom_boxplot()
```

## Check for correlation between variables
```{r correlation}

#SpCond vs:
cor.test(log_wetland$Log_SpCond_mScm, log_wetland$Log_TDS_mgl)
cor.test(log_wetland$Log_SpCond_mScm, log_wetland$Log_Sal_ppt)
cor.test(log_wetland$Log_SpCond_mScm, log_wetland$DO_mgL)
cor.test(log_wetland$Log_SpCond_mScm, log_wetland$pH)
cor.test(log_wetland$Log_SpCond_mScm, log_wetland$Log_Unfiltered_TP_ugL)
cor.test(log_wetland$Log_SpCond_mScm, log_wetland$Log_TSS_mgL)

#TDS vs:
cor.test(log_wetland$Log_TDS_mgl, log_wetland$Log_Sal_ppt)
cor.test(log_wetland$Log_TDS_mgl, log_wetland$DO_mgL)
cor.test(log_wetland$Log_TDS_mgl, log_wetland$pH)
cor.test(log_wetland$Log_TDS_mgl, log_wetland$Log_Unfiltered_TP_ugL)
cor.test(log_wetland$Log_TDS_mgl, log_wetland$Log_TSS_mgL)

#Sal vs
cor.test(log_wetland$Log_Sal_ppt, log_wetland$DO_mgL)
cor.test(log_wetland$Log_Sal_ppt, log_wetland$pH)
cor.test(log_wetland$Log_Sal_ppt, log_wetland$Log_Unfiltered_TP_ugL)
cor.test(log_wetland$Log_Sal_ppt, log_wetland$Log_TSS_mgL)

#DO vs
cor.test(log_wetland$DO_mgL, log_wetland$pH)
cor.test(log_wetland$DO_mgL, log_wetland$Log_Unfiltered_TP_ugL)
cor.test(log_wetland$DO_mgL, log_wetland$Log_TSS_mgL)

#pH vs
cor.test(log_wetland$pH, log_wetland$Log_Unfiltered_TP_ugL)
cor.test(log_wetland$pH, log_wetland$Log_TSS_mgL)

#TP vs
cor.test(log_wetland$Log_Unfiltered_TP_ugL, log_wetland$Log_TSS_mgL)

#Predictor and Outcome Variable correlations 
cor.test(log_wetland$Log_SpCond_mScm, log_wetland$Log_Filtered_NHx_ugL)
cor.test(log_wetland$Log_Cond_mScm, log_wetland$Log_Filtered_NHx_ugL)
cor.test(log_wetland$Log_TDS_mgl, log_wetland$Log_Filtered_NHx_ugL)
cor.test(log_wetland$Log_Sal_ppt, log_wetland$Log_Filtered_NHx_ugL)
cor.test(log_wetland$DO_mgL, log_wetland$Log_Filtered_NHx_ugL)
cor.test(log_wetland$pH, log_wetland$Log_Filtered_NHx_ugL)
cor.test(log_wetland$Log_Unfiltered_TP_ugL, log_wetland$Log_Filtered_NHx_ugL)
cor.test(log_wetland$Log_TSS_mgL, log_wetland$Log_Filtered_NHx_ugL)

```
SpCond and Sal are too correlated to use together - makes sense. Also all the predictor variables have very low correlations with the outcome variable :( 

# Step 3 - Fit regular regression model
I think I'm going to try three models.
```{r generalized regression models}
#remove NA's
log_wetland <- na.omit(log_wetland)

#first convert characters to factors
log_wetland$Season <- factor(log_wetland$Season, 
                           levels = c('Fall', 'Winter', 'Spring', 'Summer'))

log_wetland$Site <- as.factor(log_wetland$Site)

# fit regression models
#start with just season
mod1 <- lm(Log_Filtered_NHx_ugL ~ Season,
           data = log_wetland)

#next add predictor variables with highest correlation to NHx
mod2 <- lm(Log_Filtered_NHx_ugL ~ Season + DO_mgL + Log_Unfiltered_TP_ugL + Site,
           data = log_wetland)

#keep adding based on correlation (TSS was not significant even though it had the same value of correlation as TDS)
mod3 <- lm(Log_Filtered_NHx_ugL ~ Season + DO_mgL + Log_Unfiltered_TP_ugL +
             Log_SpCond_mScm,
           data = log_wetland)

#examine model outputs and residuals
summary(mod1)
plot(mod1)

summary(mod2)
plot(mod2)

summary(mod3)
plot(mod3)
```
* pH and TSS were not significant when I used them as predictor variables. I tried using Log_TDS but its distribution was too weird.

# Step 4 - Fit Mixed Effects Model and Examine Model Results
## use gls model first 
```{r gls}

#first refit models using generalized least squares
GLS1 <- gls(Log_Filtered_NHx_ugL ~ Season,
            data = log_wetland)

GLS2 <- gls(Log_Filtered_NHx_ugL ~ Season + DO_mgL + Log_Unfiltered_TP_ugL,
            data = log_wetland)

GLS3 <- gls(Log_Filtered_NHx_ugL ~ Season + DO_mgL + Log_Unfiltered_TP_ugL +
             Log_SpCond_mScm,
            data = log_wetland)
```


##now fit to mixed effects model and compare to GLS
```{r mixed}
#fit data to mixed effects models
MEM1 <- lme(Log_Filtered_NHx_ugL ~ Month,
            random = ~ 1|Site, data = log_wetland)
MEM2 <- lme(Log_Filtered_NHx_ugL ~ Season + DO_mgL + Log_Unfiltered_TP_ugL,
            random = ~ 1|Site, data = log_wetland)
MEM3 <- lme(Log_Filtered_NHx_ugL ~ Month + DO_mgL + Log_Unfiltered_TP_ugL +
             Log_SpCond_mScm,
            random = ~ 1|Site, data = log_wetland)

#Compare AIC values between gls and lme models
AIC(GLS1, MEM1)
AIC(GLS2, MEM2)
AIC(GLS3, MEM3)

#Examine models residual plots
plot(MEM1)
plot(MEM2)
plot(MEM3)

qqnorm(MEM1)
qqnorm(MEM2)
qqnorm(MEM3)

#Examine results
summary(MEM1)
summary(MEM2)
summary(MEM3)

```
The mixed effect models have lower AIC values than the gls linear regression models
For the mixed effect models' AIC values:
MEM2(4636) and 3(4637) are lower than MEM1(4686); MEM2 has the lowest value
MEM1 var = 0.2, MEM2 var = 0.22, MEM3 var = 0.21

I tried adding a nested model with Year/Site but it didn't improve model fit

Choosing model 2 as the best

# Step 5 - Communicate methods and results
## methods
To examine the effects of water quality parameters on ammonia/ammonium levels in NC wetlands we fit a multi-level model with filtered NHx (measured in micro grams/L) as the response variable and season, dissolved oxygen, and unfiltered total phosphorous as fixed effects. We also included site as a random effect to account for repeated sampling. 

## results
Our multi-level model results indicate that NHx concentrations significantly decreased with increasing dissolved oxygen levels (B^DO = -0.05, p < 0.001) and increased with increasing total phosphorus (B^P = 0.23, p < 0.001). There was also a significant increase in NHx concentrations in the spring (B^spring = 0.46, p <0.001), summer (B^summer = 0.50, p < 0.001), and winter (0.40, p < 0.001) relative to the fall. 

## visualize results
```{r, results}
# table results
tbl_regression(MEM2)

# plot predictions?
#Simulate DO
DO_vector <- sample(log_wetland$DO_mgL, 1460, replace = TRUE)

#Simulate TP
TP_vector <- sample(log_wetland$Log_Unfiltered_TP_ugL, 1460, replace = TRUE)

#Simulate Seasons
season_vector <- c(rep('Winter', 365),
                   rep("Spring", 365),
                   rep('Summer', 365),
                   rep('Fall', 365))

#Sites
site_vector <- rep(unique(log_wetland$Site), each = 73)

#convert chr to factors
season_vector <- factor(season_vector, 
                           levels = c('Fall', 'Winter', 'Spring', 'Summer'))

site_vector <- as.factor(site_vector)

#join with continuous data
data_pred <- data.frame(DO_vector, TP_vector, season_vector, site_vector)
colnames(data_pred) <- c('DO_mgL', 'Log_Unfiltered_TP_ugL', 'Season', 'Site')

#use original model to predict outcomes
prediction <- predict(MEM2,
                      newdata = data_pred,
                      se.fit = TRUE,
                      re.form = NA)

#revert log transformation
prediction_unlogged <- exp(prediction)

##pull out predictions
data_fig <- data.frame(data_pred,
                       prediction_unlogged)

#plot predictions using boxplot
ggplot(data_fig, aes(x = Season, y = prediction_unlogged, color = Season)) +
  geom_boxplot ()

#boxplot of raw values
ggplot(wetlands, aes(x = Season, y = Filtered_NHx_ugL, color = Season)) +
  geom_boxplot ()

#try predicted do and tp values
ggplot(data_fig, aes(x = DO_mgL, y = prediction)) +
  geom_point () +
  geom_smooth(method = "lm")

#predicted vs observed
# Create a new dataframe combining observed and predicted values
rownames(log_wetland) <- NULL
log_wetland <- log_wetland[-c(1461:1473), ]

data_fig2 <- data.frame(
  Observed_NHx = log_wetland$Log_Filtered_NHx_ugL,  
  Predicted_NHx = prediction)
  
  ggplot(data_fig2, aes(x = Observed_NHx, y = Predicted_NHx)) +
  geom_point() +                       
  geom_abline(slope = 1, intercept = 0, color = "red")


```

## just visualize raw log transformed data
```{r help}

ggplot(log_wetland, aes(x = DO_mgL, y = Log_Filtered_NHx_ugL, color = Season)) +
  geom_point () +
  facet_wrap(~Season) +
  geom_smooth(method = "lm") 

ggplot(log_wetland, aes(x = Log_Unfiltered_TP_ugL, y = Log_Filtered_NHx_ugL, color = Season)) +
  geom_point () +
  facet_wrap(~Season) +
  geom_smooth(method = "lm") 

ggplot(log_wetland, aes(y = Log_Filtered_NHx_ugL, x = Season)) +
  geom_boxplot ()

ggplot(log_wetland, aes(x = Site, y = Log_Filtered_NHx_ugL, color = Season)) +
  geom_boxplot()

```

## prediction vs observed
```{r pred}
#predict NHx values
log_wetland$predicted_nhx <- predict(MEM2)

#plot predicted vs observed NHx values
pred_vs_obs <- ggplot(log_wetland, aes(x = Log_Filtered_NHx_ugL, y = predicted_nhx)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_bw() +
  labs(title = "Predicted vs Observed NHₓ",
       x = "Observed Log(NHₓ µg/L)",
       y = "Predicted Log(NHₓ µg/L)") +
  theme(axis.text = element_text(size = 14, color = "black"),
        axis.title = element_text(size = 14),
        title = element_text(size = 16))

#save image
ggsave("NHx_pred_vs_obs.png", pred_vs_obs, width = 12, height = 6, dpi = 300)


residuals <- resid(MEM2)
fitted <- fitted(MEM2)

res_df <- data.frame(
  Fitted = fitted,
  Residuals = residuals
)

 res_vs_fit <- ggplot(res_df, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_bw() +
  labs(title = "Residuals vs. Fitted",
       x = "Fitted Values",
       y = "Residuals") +
  theme(axis.text = element_text(size = 14, color = "black"),
        axis.title = element_text(size = 14),
        title = element_text(size = 16))
 
#NHx_diagnostics <- plot_grid(pred_vs_obs, res_vs_fit, labels = c("A", "B"), ncol = 2, align = "h")

#ggsave("NHx_diagnostics.png", NHx_diagnostics, width = 12, height = 6, dpi = 300)

```
