---
title: "NOx_models_NG"
output: pdf_document
date: "2025-04-02"
---


# Libraries and Data Loading

```{r}

library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(moments)
library(readxl)
library(lubridate)
library(lme4)
library(nlme)
library(gtsummary)
library(cowplot)


wetlands <- read.csv(here('Data/Processed/Log_Transformed_Data.csv'))

wetlands <- na.omit(wetlands)

glimpse(wetlands)
colnames(wetlands)

```


# Data Exploration: Continuous Variables Histograms

```{r}

ggplot(wetlands, aes(x = Log_SpCond_mScm)) +
geom_histogram()

skewness(wetlands$Log_SpCond_mScm, na.rm = T)
kurtosis(wetlands$Log_SpCond_mScm, na.rm = T)

ggplot(wetlands, aes(x = Log_Cond_mScm)) +
geom_histogram()

skewness(wetlands$Log_Cond_mScm, na.rm = T)
kurtosis(wetlands$Log_Cond_mScm, na.rm = T)

ggplot(wetlands, aes(x = Log_TDS_mgl)) +
geom_histogram()

skewness(wetlands$Log_TDS_mgl, na.rm = T)
kurtosis(wetlands$Log_TDS_mgl, na.rm = T)

ggplot(wetlands, aes(x = Log_Sal_ppt)) +
geom_histogram()

skewness(wetlands$Log_Sal_ppt, na.rm = T)
kurtosis(wetlands$Log_Sal_ppt, na.rm = T)

ggplot(wetlands, aes(x = DO_percent)) +
geom_histogram()

skewness(wetlands$DO_percent, na.rm = T)
kurtosis(wetlands$DO_percent, na.rm = T)

ggplot(wetlands, aes(x = DO_mgL)) +
geom_histogram()

skewness(wetlands$DO_mgL, na.rm = T)
kurtosis(wetlands$DO_mgL, na.rm = T)

ggplot(wetlands, aes(x = pH)) +
geom_histogram()

skewness(wetlands$pH, na.rm = T)
kurtosis(wetlands$pH, na.rm = T)

ggplot(wetlands, aes(x = Log_TSS_mgL)) +
geom_histogram()

skewness(wetlands$Log_TSS_mgL, na.rm = T)
kurtosis(wetlands$Log_TSS_mgL, na.rm = T)

ggplot(wetlands, aes(x = fecal_coliform_colonies_per100mL)) +
geom_histogram()

skewness(wetlands$fecal_coliform_colonies_per100mL, na.rm = T)
kurtosis(wetlands$fecal_coliform_colonies_per100mL, na.rm = T)

ggplot(wetlands, aes(x = Log_Filtered_NOx_ugL)) +
geom_histogram()

skewness(wetlands$Log_Filtered_NOx_ugL, na.rm = T)
kurtosis(wetlands$Log_Filtered_NOx_ugL, na.rm = T)



```


From the histogram plots, and skewness and kurtosis values, the variables that are normally distributed enough for analysis are DO%, DO (mg/L), pH, and log(TSS). Assessments of categorical variables will follow: 


# Data Exploration: Categorical Variables Bar Charts

```{r}

unique(wetlands$Site)

ggplot(wetlands, aes(Site))+
  geom_bar()

ggplot(wetlands, aes(Month))+
  geom_bar()

ggplot(wetlands, aes(Year))+
  geom_bar()

ggplot(wetlands, aes(Season))+
  geom_bar()


```


# Data Exploration: Scatter and Box Plots

```{r}
#Continuous Variables: 
#- DO% / DO mg_L/ pH / LogTSS
#Categorical Variables:
#- Site/ month/ Year/ season

DO_types_plot <- ggplot(wetlands, aes(x = DO_mgL, y = DO_percent)) +
  geom_point()
DO_types_plot

#variables extremely correlated (as expected) should not be used in the same model

DO_pH_plot <- ggplot(wetlands, aes(x = pH, y = DO_percent)) +
  geom_point()
DO_pH_plot

  
DO_pH_plot2 <- ggplot(wetlands, aes(x = pH, y = DO_mgL)) +
  geom_point()
DO_pH_plot2

#pH does not seem correlated with either DO mgl or DO %

DO_TSS_plot <- ggplot(wetlands, aes(x = Log_TSS_mgL, y = DO_percent)) +
  geom_point()
DO_TSS_plot

  
DO_TSS_plot2 <- ggplot(wetlands, aes(x = Log_TSS_mgL, y = DO_mgL)) +
  geom_point()
DO_TSS_plot2

#logTSS does not seem correlated with either DO mgl or DO %

DO_season_plot <- ggplot(wetlands, aes(x = Season, y = DO_mgL)) +
  geom_boxplot()
DO_season_plot

#season appears to influence DO mgL but not too much to avoid using together


TSS_season_plot <- ggplot(wetlands, aes(x = Season, y = Log_TSS_mgL)) +
  geom_boxplot()
TSS_season_plot

#season does not appear to strongly influence TSS 

#site based differences?

TSS_site_plot <- ggplot(wetlands, aes(x = Site, y = Log_TSS_mgL)) +
  geom_boxplot()
TSS_site_plot

#potential influences of site on TSS

O2_site_plot <- ggplot(wetlands, aes(x = Site, y = DO_mgL)) +
  geom_boxplot()
O2_site_plot

#no signif differences


```


# Linear Models


## Which Oxygen Measurement to Use

```{r}

#Determining which oxygen measurement would be best to use in the models
wetlandsO2_perc.lm <- lm(Log_Filtered_NOx_ugL ~ DO_percent,
                       data = wetlands)
summary(wetlandsO2_perc.lm)
AIC(wetlandsO2_perc.lm)
#DO % AIC = 5991.301

wetlandsO2_mgL.lm <-  lm(Log_Filtered_NOx_ugL ~ DO_mgL,
                         data = wetlands)
summary(wetlandsO2_mgL.lm)
AIC(wetlandsO2_mgL.lm)
#DO mgL AIC = 6010.893

#The oxygen measurements were fairly similar in predictive power, however the DO% had a slightly more powerful AIC value and will be used in further models




```

# further modeling

```{r}

#fix AIC and r squared

seasonbased.lm <- lm(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + pH + Season + Site + Year,
                       data = wetlands)
summary(seasonbased.lm)
AIC(seasonbased.lm)
#AIC: 4588.897
#Adj R2: 0.5487
#p-value: < 2.2e-16


#this model provides the best fit for predicting NO3 concentrations
monthbased.lm <- lm(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + pH + Month + Site + Year,
                       data = wetlands)
summary(monthbased.lm)
AIC(monthbased.lm)
#AIC: 4563.233
#adj r2: 0.5559 
#p-value: < 2.2e-16


#removed pH/ year because was the least significant, kept month over season because lower AIC value/ higher adj r2
nopH.lm <- lm(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + Month + Site,
                       data = wetlands)
summary(nopH.lm)
AIC(nopH.lm)
plot(nopH.lm)
#AIC: 4560.75
#multiple r2: 0.556  
#p-value: < 2.2e-16

AIC(seasonbased.lm, monthbased.lm, nopH.lm)

## AIC/ R2 is only slightly better, however we are going to go with the simpler model?

```


Lm with no pH/ year had the lowest degrees of freedom/ was the simplest model/ had the lowest AIC - will be used for nested/ mixed effects

##Mixed Effects Models

```{r}


nox.lmm1 <- lme(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + Month, 
                  random = ~1|Site, data = wetlands)

plot(nox.lmm1)
qqnorm(nox.lmm1)
summary(nox.lmm1)
AIC(nox.lmm1)

nox.lmm2 <- lme(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + Month, 
                  random = ~1|Year, data = wetlands)

plot(nox.lmm2)
qqnorm(nox.lmm2)
summary(nox.lmm2)
AIC(nox.lmm2)

nox.lmm3 <- lme(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + Month, 
                  random = ~1|Site/Year, data = wetlands)

plot(nox.lmm3)
qqnorm(nox.lmm3)
summary(nox.lmm3)
AIC(nox.lmm3)
tbl_regression(nox.lmm3)

AIC(nox.lmm1, nox.lmm2, nox.lmm3)


#refit model 
# even with refitting, month provided a better AIC than 
monthbased.gls <- gls(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + Month + Site,
                       data = wetlands)
summary(monthbased.gls)
AIC(monthbased.gls)
#AIC: 4628.224

seasonbased.gls <- gls(Log_Filtered_NOx_ugL ~ DO_percent + Log_TSS_mgL + Season + Site,
                       data = wetlands)
summary(monthbased.gls)
AIC(seasonbased.gls)
#AIC: 4657.308

AIC(monthbased.gls, seasonbased.gls)


AIC(monthbased.gls, nox.lmm3)


```


Best model: mixed effects model with site nested in year. Interesting because initially took year out of the linear model. 


##plots

```{r}

#predicted vs observed plots

wetlands$predicted_nox <- predict(nox.lmm3)

pred_obs_NOx <- ggplot(wetlands, aes(x = Log_Filtered_NOx_ugL, y = predicted_nox)) +
  geom_point(alpha = 0.4) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
  theme_bw() +
  labs(title = "Predicted vs Observed NOₓ",
       x = "Observed Log(NOₓ µg/L)",
       y = "Predicted Log(NOₓ µg/L)") +
  theme(axis.text = element_text(size = 14, color = "black"),
        axis.title = element_text(size = 14),
        title = element_text(size = 16))
pred_obs_NOx

#Residuals plot
residuals <- resid(nox.lmm3)
fitted <- fitted(nox.lmm3)

res_df <- data.frame(
  Fitted = fitted,
  Residuals = residuals
)

residvfit <- ggplot(res_df, aes(x = Fitted, y = Residuals)) +
  geom_point(alpha = 0.4) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  theme_bw() +
  labs(title = "Residuals vs. Fitted",
       x = "Fitted Values",
       y = "Residuals") +
  theme(axis.text = element_text(size = 14, color = "black"),
        axis.title = element_text(size = 14),
        title = element_text(size = 16))

NOx_diagnostics <- plot_grid(pred_obs_NOx, residvfit, labels = c("A", "B"), ncol = 2, align = "h")

ggsave("NOx_diagnostics.png", NOx_diagnostics, width = 12, height = 6, dpi = 300)


```





