---
title: "NOx_models_NG"
output: pdf_document
date: "2025-04-02"
---


## Libraries and Data Loading

```{r}

library(here)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(moments)
library(readxl)
library(lubridate)


wetlands <- read.csv(here('Data/Processed/Log_Transformed_Data.csv'))

glimpse(wetlands)
colnames(wetlands)

```


## Data Exploration: Continuous Variables

```{r}


ggplot(wetlands, aes(x = Log_SpCond_mScm)) +
geom_histogram()

skewness(wetlands$Log_SpCond_mScm, na.rm = T)
kurtosis(wetlands$Log_SpCond_mScm, na.rm = T)

ggplot(wetlands, aes(x = Log_Cond_mScm)) +
geom_histogram()

skewness(wetlands$Log_Cond_mScm, na.rm = T)
kurtosis(wetlands$Log_Cond_mScm, na.rm = T)

ggplot(wetlands, aes(x = Log_TDS_mgl)) +
geom_histogram()

skewness(wetlands$Log_TDS_mgl, na.rm = T)
kurtosis(wetlands$Log_TDS_mgl, na.rm = T)

ggplot(wetlands, aes(x = Log_Sal_ppt)) +
geom_histogram()

skewness(wetlands$Log_Sal_ppt, na.rm = T)
kurtosis(wetlands$Log_Sal_ppt, na.rm = T)

ggplot(wetlands, aes(x = DO_percent)) +
geom_histogram()

skewness(wetlands$DO_percent, na.rm = T)
kurtosis(wetlands$DO_percent, na.rm = T)

ggplot(wetlands, aes(x = DO_mgL)) +
geom_histogram()

skewness(wetlands$DO_mgL, na.rm = T)
kurtosis(wetlands$DO_mgL, na.rm = T)

ggplot(wetlands, aes(x = pH)) +
geom_histogram()

skewness(wetlands$pH, na.rm = T)
kurtosis(wetlands$pH, na.rm = T)

ggplot(wetlands, aes(x = Log_TSS_mgL)) +
geom_histogram()

skewness(wetlands$Log_TSS_mgL, na.rm = T)
kurtosis(wetlands$Log_TSS_mgL, na.rm = T)

ggplot(wetlands, aes(x = fecal_coliform_colonies_per100mL)) +
geom_histogram()

skewness(wetlands$fecal_coliform_colonies_per100mL, na.rm = T)
kurtosis(wetlands$fecal_coliform_colonies_per100mL, na.rm = T)

ggplot(wetlands, aes(x = Log_Filtered_NOx_ugL)) +
geom_histogram()

skewness(wetlands$Log_Filtered_NOx_ugL, na.rm = T)
kurtosis(wetlands$Log_Filtered_NOx_ugL, na.rm = T)



```


From the histogram plots, and skewness and kurtosis values, the variables that are normally distributed enough for analysis are DO%, DO (mg/L), pH, and log(TSS). Assessments of categorical variables will follow: 


## Data Exploration: Categorical Variables

```{r}

unique(wetlands$Site)

ggplot(wetlands, aes(Site))+
  geom_bar()

ggplot(wetlands, aes(Month))+
  geom_bar()

ggplot(wetlands, aes(Year))+
  geom_bar()

ggplot(wetlands, aes(Season))+
  geom_bar()


##filter out first and last years
##filter out AN3/ AS/ SP/TV/ UP

wetlands2 <- wetlands %>% 
  filter(!(Year  %in% c(2015, 2025))) %>% 
  filter(!(Site %in% c("AN3", "AS", "SP", "TV", "UP", "DP")))

ggplot(wetlands2, aes(Site))+
  geom_bar()

ggplot(wetlands2, aes(Year))+
  geom_bar()


```



## NEED TO do correlations first!!

## Linear Models

```{r}

wetlandsO2_perc.lm <- (Log_Filtered_NOx_ugL ~ DO_percent + pH + Log_TSS_mgl + Season)
  
wetlandsO2_mgL.lm <- 


```



Notes: 

Continuous Variables: 
- DO% / DO mg_L/ pH / LogTSS
Categorical Variables:
- Site/ month/ Year/ season

Look at
- DO% vs DO mg/L
- season v month v year
- combinations of season + year/ month + year

Random Effects
- site
- year
- potentially site nested in year

